---
title: "SimiCviz: Visualization of SimiC Outputs"
author: "Irene Marín-Goñi"
output:
  BiocStyle::html_document:
    toc: true
vignette: >
  %\VignetteIndexEntry{SimiCviz: Visualization of SimiC Outputs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  eval = TRUE,
  comment = "##"
)
options(width=60)
# install.packages("magick")
# library(magick)
library(SimiCviz)
```

# Overview

This vignette demonstrates how to use **SimiCviz** to visualize SimiCPipeline outputs and other gene regulatory network outputs.

# Input data

## SimiCPipeline outputs
You can read standard SimiC results from pickle and CSV files. 
If you followed the tutorials from [`SimiCPipeline`](https://github.com/ML4BM-Lab/SimiCPipeline.git) you will have a directory set like:
```
Project/
├── inputFiles/
│   ├── TF_list.csv
│   ├── expression_matrix.pickle
│   └── phenotype_annotation.txt
└── outputSimic/
    ├── figures/
    └── matrices/
        └── example1/
            ├── example1_L1_0.1_L2_0.01_simic_matrices.pickle
            ├── example1_L1_0.1_L2_0.01_simic_matrices_filtered_BIC.pickle
            ├── example1_L1_0.1_L2_0.01_wAUC_matrices_filtered_BIC.pickle
            └── example1_L1_0.1_L2_0.01_wAUC_matrices_filtered_BIC_collected.csv

```
### Standard Output Directory
SimiCviz provides the following functions to import SimiCPipeline outputs easily.
```{r collapse=TRUE, results='hold'}
library(SimiCviz)
simic <- load_SimiCPipeline(project_dir = "/home/workdir/SimiCviz/inst/extdata",
                               run_name = "example1",
                               lambda1 = 0.01,
                               lambda2 = 0.001)

message("\nSetting lables and colors")
simic <- setLabelNames(simic, label_names  = c("control","PD-L1","DAC","Combo"),
              colors = c("#e0e0e0", "#a8c8ff", "#ffb6b6", "#c1a9e0"))
simic

```

### Manual Loading
You can load the different simic outputs and construct your own object

```{r}
weights_file <- system.file("extdata",file.path("outputSimic/matrices/example1",
                  "example1_L1_0.01_L2_0.001_simic_matrices_filtered_BIC.pickle"), 
                            package = "SimiCviz")

out <- read_pickle(weights_file)
str(out,1)
weights_results <- read_weights_pickle(weights_file)
head(weights_results[[1]][,1:6])
```
### Activity scores

The csv generated by SimiCPipeline contains the activity scores for all the TFs in the experiment and all the cells in the input matrix.

```{r, eval= FALSE}
auc_collect_file <- system.file("extdata",file.path("outputSimic/matrices/example1"
                                ,"example1_L1_0.01_L2_0.001_wAUC_matrices_filtered_BIC_collected.csv"),
                                package = "SimiCviz")
auc_collect <- read.csv(auc_collect_file, header=TRUE, row.names = 1)
head(auc_collect[,1:5])

```

If you want to access to the raw activity scores matrices you can use the read_auc_csv function
```{r}
auc_file <- system.file("extdata", file.path("outputSimic/matrices/example1"
                        ,"example1_L1_0.01_L2_0.001_wAUC_matrices_filtered_BIC.pickle"), 
                            package = "SimiCviz")
auc_pkl <- read_auc_pickle(auc_file)

print(class(auc_pkl))
print(length(auc_pkl))
```

For full plotting functionalities you will need to provide the cell labels id.

## Load SimiCvizExperiment
With these `SimiCPipeline` output we can generate the `SimiCvizExperiment` object easily.
```{r}
cell_labels_path <- system.file("extdata",file.path("inputFiles","treatment_annotation.csv"), package = "SimiCviz")
```


```{r}
# all(out$TF_ids%in%colnames(auc_pkl[[1]])) # TRUE
# all(out$query_targets%in%colnames(weights_results[[1]])) # TRUE

simic2 <- SimiCvizExperiment(weights = weights_results,
                             auc = auc_pkl,
                             cell_labels = cell_labels_path,
                             tf_ids = out$TF_ids,
                             target_ids = out$query_targets,
                             label_names = c("control","PD-L1","DAC","Combo"),
                             colors = c("#e0e0e0", "#a8c8ff", "#ffb6b6", "#c1a9e0"))
simic2
```

### Other data structures

Alternatively you can load the results from other methods in csv format. 

SimiCviz expects for the weights a csv with at least columns `c("tf","target","weight")` and for the activity scores a csv with columns `c("cell", "tf","score")`
```{r  collapse=TRUE}
weight_path <-  system.file("extdata","example_weights.csv", 
                            package = "SimiCviz")
weight_df <- read_weights_csv(weight_path)
head(weight_df)

auc_path <-  system.file("extdata","example_auc.csv", package = "SimiCviz")
auc_df <- read_auc_csv(auc_path)
head(auc_df)

# Loading using the paths
simic3 <- load_from_csv(weights_file = weight_path, 
                                #  auc = auc_path, 
                                 meta = list(lambda1 = 0.01, lambda2= 0.001))
simic3 <- setLabelNames(simic3, label_names  = c("control","PD-L1","DAC","Combo"),
                        colors = c("#e0e0e0", "#a8c8ff", "#ffb6b6", "#c1a9e0"))

# Loading using the dataframe (equivalent)
# simic3 <- SimiCvizExperiment(weights = weight_df,
#                              auc = auc_df,
#                              tf_ids = unique(auc_df$tf),
#                              target_ids = unique(weight_df$target),
#                              meta = list(lambda1 = 0.01, lambda2= 0.001),
#                              label_names = c("control","PD-L1","DAC","Combo"),
#                              colors = c("#e0e0e0", "#a8c8ff", "#ffb6b6", "#c1a9e0"))
simic3
```
# Weight Visualizations
```{r}
plot_dir <- file.path(getwd(),"SimiCviz_output","plots")
dir.create(plot_dir,recursive = TRUE)
```
## Weight R2 assessment
```{r }
adjusted_r_squared <- simic@meta$adjusted_r_squared
plot_r2_distribution(adjusted_r_squared, simic, grid=c(2,2), width= 16,save =TRUE, out_dir = plot_dir)
```

```{r, fig.height=8, fig.width=10,echo =FALSE, message = FALSE}
plot_r2_distribution(adjusted_r_squared, simic, grid=c(2,2))
```

## TF Weight barplots

```{r}
plot_tf_weights(simic, 
                tf_names = simic@tf_ids[1:4],
                top_n    = 30, 
                save = TRUE, 
                grid     = c(4, 1),
                out_dir = plot_dir, 
                filename = "TF_weights_barplot.pdf")
```

```{r, fig.height=8, fig.width=10, echo = FALSE, message = FALSE}
plot_tf_weights(simic, 
                tf_names = simic@tf_ids[1:4],
                top_n    = 30, 
                save = FALSE, 
                grid     = c(4, 1))
```

## Target Weight barplots

```{r, eval = TRUE}
plot_target_weights(simic, 
                target_names = simic@target_ids[1:4],
                labels   = c("control","DAC"),
                save = TRUE, 
                width= 18, # Adjust width and height as needed for grid layout
                height = 10, 
                grid     = c(2, 2),
                out_dir = plot_dir, 
                filename = "Target_weights_barplot.pdf")
```

```{r,fig.height=10, fig.width=15, echo = FALSE, message = FALSE}
plot_target_weights(simic, 
                target_names = simic@target_ids[1:4],
                labels   = c("control","DAC"),
                save = FALSE, 
                width= 18, height = 10,
                grid     = c(2, 2))
```


If you want to plot all TFs and Targets bar plots leave the `tf_names` and `target_names` arguments respectivelly blank.

```{r, eval = FALSE}
plot_tf_weights(simic, 
                save = TRUE, 
                width= 18, height = 10,
                grid     = c(3, 2),
                out_dir = plot_dir,
                filename = "All_tf_weights_barplot.pdf")
plot_target_weights(simic, 
                save = TRUE, 
                width= 18, height = 10,
                grid     = c(3, 2),
                out_dir = plot_dir,
                filename = "All_target_weights_barplot.pdf")
```

## Network visualization

```{r, fig.height=8, fig.width=10, eval = FALSE}
plot_network(simic, top_n = 4)
```
# AUC Visualizations
## Density distributions
Plot per-TF activity score density curves across phenotype labels.
```{r message=FALSE}
# Plot top 4 TFs density distributions
plot_auc_distributions(simic,
                       tf_names = simic@tf_ids[1:4],
                       fill = TRUE,
                       alpha = .6,
                       bw_adjust = 1/8,
                       rug = TRUE,
                       save = TRUE,
                       out_dir = plot_dir,
                       filename = "AUC_distributions.pdf",
                       grid = c(2, 2))

```
```{r echo = FALSE ,fig.height=10, fig.width=15}
# Plot top 4 TFs density distributions
plot_auc_distributions(simic,
                       tf_names = simic@tf_ids[1:4],
                       fill = TRUE,
                       alpha = 0.6,
                       bw_adjust = 1/8,
                       rug = TRUE,
                       save = FALSE,
                       grid = c(2, 2))
```
```{r message=FALSE ,fig.height=10, fig.width=15}
# Plot top 6 TFs density distributions
plot_auc_distributions(simic,
                       labels = c(0,2,3),
                       tf_names = simic@tf_ids[1:6],
                       fill = TRUE,
                       alpha = 0.9,
                       bw_adjust = 0.8,
                       rug = TRUE,
                       out_dir = plot_dir,
                       filename="AUC_distributions_filled.pdf",
                       save = TRUE,
                       grid = c(2, 3))
```
```{r echo = FALSE ,fig.height=10, fig.width=15}
# Plot top 6 TFs density distributions
plot_auc_distributions(simic,
                       labels = c(0,2,3),
                       tf_names = simic@tf_ids[1:6],
                       fill = TRUE,
                       alpha = 0.9,
                       bw_adjust = 0.8,
                       rug = TRUE,
                       save = FALSE,
                       grid = c(2, 3))
```
```{r  }
# Plot top 4 TFs density distributions
plot_auc_distributions(simic,
                       labels = c(0,3),
                       tf_names = simic@tf_ids[1:2],
                       fill = FALSE,
                       alpha = 0.9,
                       bw_adjust = 0.5,
                       rug = TRUE,
                       out_dir = plot_dir,
                       filename="AUC_distributions_notfilled_multipage.pdf",
                       save = TRUE,
                       grid = c(1,2))
```
```{r echo = FALSE ,fig.height=10, fig.width=15}
# Plot top 4 TFs density distributions
plot_auc_distributions(simic,
                       labels = c(0,3),
                       tf_names = simic@tf_ids[1:2],
                       fill = FALSE,
                       alpha = 0.9,
                       bw_adjust = 0.5,
                       rug = TRUE,
                       save = FALSE,
                       grid = c(2, 2))
```

## Cumulative distributions (ECDF)
```{r message=FALSE, warning = FALSE}
plot_auc_cumulative(simic,
                    tf_names = simic@tf_ids[1:2],
                    rug = FALSE,
                    grid = c(2, 2),
                    save = TRUE, 
                    include_table = TRUE,
                    width = 12, height = NULL,
                    out_dir = plot_dir)
```
```{r echo = FALSE, fig.height=8, fig.width=15}
plot_auc_cumulative(simic,
                    tf_names = simic@tf_ids[1:2],
                    rug = FALSE,
                    grid = c(1, 2),
                    save = FALSE)
```

# Additional functions

You can convert the activity scores file into a data.frame of cells x TFs with the following function:
```{r eval = FALSE}
auc_file <- system.file("extdata", file.path("outputSimic/matrices/example1"
                        ,"example1_L1_0.01_L2_0.001_wAUC_matrices_filtered_BIC.pickle"), 
                            package = "SimiCviz")
auc_pkl <- read_auc_pickle(auc_file)

cell_labels_path <- system.file("extdata",file.path("inputFiles","treatment_annotation.csv"), package = "SimiCviz")
cell_labels = load_cell_labels(cell_labels_path)
auc_df <- auc_list_to_df(auc_pkl, cell_labels_df = cell_labels )
head(auc_df[,1:5])

save_collected_auc(auc_df, file = "./collected_auc.csv", overwrite = FALSE)
```






## ECDF-AUC metrics table

```{r eval = FALSE}
ecdf_metrics <- calculate_ecdf_auc(simic, tf_names = simic@tf_ids[1:6])
head(ecdf_metrics)
```
## Summary statistics (boxplot, violin, mean bar, high-activity count)

```{r eval = FALSE}
plot_auc_summary_statistics(simic)
```

## AUC Heatmap (mean activity per TF × phenotype)

```{r, fig.height=8, fig.width=10, eval=FALSE}
plot_auc_heatmap(simic, top_n = 20)
```

If AUC data is condition-specific, you can use a heatmap:

```{r, eval=FALSE}
plot_auc_heatmap(simic)
```

# Exporting results

```{r, eval=FALSE}
# getwd()
# out_dir <- "/home/workdir/SimiCviz/SimiCviz_output"
# dir.create(file.path(out_dir))
# export_SimiCviz_all(simic, out_dir = out_dir, prefix = "SimiC_example")
```

This will create:

- `data/weights`, `data/auc`
- `plots/network`, `plots/auc`

under `SimiCviz_output/`.
